---
title: "Visualizing data on the fastest growing companies"
author: "Heather Geiger"
date: February 10, 2019
output:
  html_document:
   code_folding: hide
  html_notebook: default
---

**Principles of Data Visualization and Introduction to ggplot2**

I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}

inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header=TRUE)

```

And lets preview this data:

```{r}

head(inc)

```
```{r}

summary(inc)

```

Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

```{r}
print("Top 10 growth rates:")
head(inc$Growth_Rate[order(inc$Growth_Rate,decreasing=TRUE)],n=10)

print("Top 10 revenue values, in millions:")
head(inc$Revenue[order(inc$Revenue,decreasing=TRUE)]/1e6,n=10)

print("Top 10 numbers of employees:")
head(inc$Employees[order(inc$Employees,decreasing=TRUE)],n=10)

print("Number of industries in this data set:")
length(unique(as.vector(inc$Industry)))

frequency_per_industry <- data.frame(table(as.vector(inc$Industry)))
frequency_per_industry <- frequency_per_industry[order(frequency_per_industry$Freq,decreasing=TRUE),]

print("Full frequency table of industries vs. # of companies:")
frequency_per_industry

print("Number of cities in this data set:")
length(unique(as.vector(inc$City)))
```

First of all, we see that the Rank column is just going from 1 to 5000, ranking in descending order by growth rate.

The name column just gives the name of each company.

12 companies are missing information on the number of employees.

I noticed that Growth_Rate, Revenue, and Employees all had a max value much larger than the third quartile value. This made me wonder if these max values were just outliers, or if the distributions were right-skewed. All three of these variables ended up having at least 10 values much higher than the third quartile value. This suggests that the variables are probably right-skewed, even before we visualize the distribution using a histogram.

I showed the full distribution for number of companies per industry, since there were not that many different industries in the data set. This data would be best represented in a barplot, but we can get an idea just looking at the table. Certain industries (such as IT Services and Business Products & Services) have a lot more companies in this list than others.

Finally, we find that certain cities and states also have a lot more companies on the list. Given 50 states, one state having 300+ companies is not expected, and could be explained by factors such as population size or economic policies. Similarly there are over 1500 cities on the list. So this means there are a few cities with a lot more companies, and then a lot of cities that have only one or two companies.

## Question 1

Create a graph that shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.

```{r,fig.asp = 1.5}

library(ggplot2)

inc <- within(inc,
	State <- factor(State,
		levels=names(sort(table(State)))))

ggplot(inc,
aes(x=State)) +
geom_bar() +
coord_flip() +
ylab("Number of companies") +
ggtitle("Fastest growing companies, by state")

```

## Question 2

Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

```{r, fig.asp=1.5}
nyinc <- inc[as.vector(inc$State) == "NY",]
nyinc <- nyinc[complete.cases(nyinc),]

median_employment_per_industry <- aggregate(Employees ~ Industry,data=nyinc,FUN=median)
median_employment_per_industry <- median_employment_per_industry[order(median_employment_per_industry$Employees),]

nyinc$Industry <- factor(nyinc$Industry,
	levels=as.vector(median_employment_per_industry$Industry))

ggplot(nyinc,
aes(x=Industry,y=Employees)) +
geom_boxplot() +
coord_flip(ylim=c(0,1500)) +
stat_summary(fun.y=mean, colour="darkred", geom="point",
	shape=18, size=3) +
ylab("Employees per company") +
ggtitle("NY State companies on list of 5,000 fastest growing in US\nEmployment in these companies, by industry") 
```

In this plot, I deal with outliers by:

1. I use ylim inside of coord_flip to show only from the part of the plot from 0 to 1500 employees. The values for >1500 employees are still in the data, just not showing that part of the plot.
2. Display the mean (red diamonds) in addition to the median. In cases where the mean is much larger than the median, one can infer the presence of outliers even if these outliers are not visible within the plot area (mainly relevant for Business Products and Services and Consumer Products and Services).

Conclusions from this plot (and what we can't learn just from this one plot):

1. Many industries have distributions that are right-skewed and/or have outliers. This means most companies in the industry have relatively few employees, while there are one or a few that are much much larger.
2. Just based on this plot, environmental services and energy stand out as having a substantially higher median number of employees per company. E.g. if you pick out any given company in these industries, they will tend to employ a lot of residents in the state. Although what this doesn't show is how many companies in each industry there are. If it turns out that there are only two environmental services companies in the whole state, this may not be a super exciting industry even if each of those companies employs 100+ people.

## Question 3

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

```{r,fig.asp=1.5}

inc <- inc[complete.cases(inc),]

median_employment_per_industry <- aggregate((Revenue/Employees) ~ Industry,data=inc,FUN=median)
median_employment_per_industry <- median_employment_per_industry[order(median_employment_per_industry[,2]),]

inc$Industry <- factor(inc$Industry,
        levels=as.vector(median_employment_per_industry$Industry))

ggplot(inc,
aes(x=Industry,y=(Revenue/1000000)/Employees)) +
geom_boxplot() +
coord_flip(ylim=c(0,2)) +
stat_summary(fun.y=mean, colour="darkred", geom="point",
	shape=18, size=3) +
ylab("Companies' average revenue per employee, $millions") +
ggtitle("Top 5,000 fastest growing companies in US")

```

Conclusions from this plot (and what we can't learn just from this one plot):

1. Once again many industries have distributions that are right-skewed and/or have outliers. This means most companies in the industry have relatively low average revenue per employee, while there are one or a few more companies with a lot higher revenue ratio relative to employees.
2. Once again, this plot does not include information on how many companies from each industry are on the list. For example, this plot makes computer hardware seem like a really promising industry to invest in. But it does not tell you what we saw in our initial exploration, that less than 1% of companies on the list are in this industry. So you might not get a lot of opportunities to invest in computer hardware companies.
